Hierarchical wrinkling in a confined permeable biogel

### 2015-02-18 fig_plis/article_plis.tex
Petites corrections de la biblio et des SI par Seb

\paragraph*{Acknowledgements}
Confocal experiments were conducted at SFR BioSciences Gerland - Lyon Sud (US8 / UMS3444).
The authors thank Jos\'e Bico, Arezeki Boudaoud, Cyprien Gay, L. Mahadevan and Olivier Pierre-Louis for theoretical insights at various stages of the research. 
Precious experimental ideas came from Thibaut Divoux, Catherine Barentin, Denis Bartolo, Elisabeth Bouchaud and Maxime Lefranc.
Alan Parker at Firmenich provided the casein and GDL; 
Special thanks to Madame Biot who made M.A. Biot work accessible through the \href{http://www.olemiss.edu/sciencenet/poronet/}{poronet} website. 
ML thanks the Region Rhône Alpes and the Programme d'Avenir Lyon - Saint Etienne (PALSE NoGELPo) for postoctoral grant. 
TG and MN acknowledge funding from the agence national de la recherche fran\c{c}aise (ANR-11-PDOC-027). 
SM and ML acknowledge funding from the European Research Council under the European Union's Seventh Framework Program (FP7/2007-2013) / ERC grant agreement No. 258803.

@@

\subsection*{Fraction of free casein as a function of the pH}

The fraction of free casein proteins of a \SI{100}{\milli\litre}, 4\%w sodium caseinate dispersion is obtained either by titration by HCl 1 molar or during the action of 4\%w GDL. In both cases, \SI{0.4}{\milli\litre} of the \SI{100}{\milli\litre} dispersion is sampled regularly as a function of the pH, and centrifuged (centrifuge 5424, Eppendorf) in Eppendorf tubes at 14500 rpm for \SI{3}{\minute}. The supernatant is diluted and analysed with a UV absorption spectrometer (USB4000, Ocean Optics). According to~\cite{Roefs1986}, we deduced the fraction of free casein proteins from the adsorption spectrum:
%
\begin{align}
x_{free} &= \frac{A(\text{pH})}{A(\text{pH}=7)},\\
\text{with }A(\text{pH}) &= A_{280}-1.706 A_{320},
\end{align}
%
where $A_{280}$ is the absorption peak due to aromatic amino acids in the casein molecules plus a turbidity background and $A_{320}$ is only due to turbidity.

Above the isoelectric point, the homogeneous acidification by GDL continuously creates very small protein aggregates that are difficult to centrifuge. Such suspensions are too turbid and too rapidly evolving to reliably measure $x_{free}$.

@@

\begin{figure}
	\includegraphics{nonmonotonic}%
	\caption{Microstructure evolution. (a) Details of confocal images in fully adhesive conditions (top row) and with no adhesion on top (bottom row). The scale bar is \SI{10}{\micro\metre}. Arrows indicates the pore size. (b) Corresponding Fourier spectra when no adhesion on top, same times as (a). Lines are fits by equation~(\ref{eq:fractalS}). The inset shows a schematic association between real space structures and wave numbers. (c-e) Evolution of cut-off length, susceptibility and fractal dimension in full adhesive (black line) and no adhesion on top (orange dashed line) situations. Symbols indicates the times of the images in (a).}%
	\label{fig:nonmonotonic}%
\end{figure}

To follow the evolution of the microstructure of the gel, we acquire slices of $(\SI{2048}{px})^2$ with pixels of \SI{98}{\nano\metre}, Supplementary Figure~\ref{fig:nonmonotonic}a and Supplementary Videos~\ref{vid:stick63} and \ref{vid:nostick63}. After Hann windowing we perform Fourier transform and radially average the resulting spectrum to obtain $I(q)$, see Supplementary Figure~\ref{fig:nonmonotonic}b that is fitted by the following fractal form:
\begin{equation}
I(q) = \frac{2\chi\Gamma(d)}{\left(1+\left((d+1)\xi q\right)^2\right)^{d/2}}
\label{eq:fractalS}
\end{equation}
where the susceptibility $\chi$ is related to the fluorescence contrast between the network and the solvent, the cut off length $\xi$ corresponds to the largest pore size and $d$ is the fractal dimension. $\Gamma$ is the gamma function. Equation~(\ref{eq:fractalS}) is a good approximation to the 2D Fourier transform of a fractal pair correlation function $g(r) \sim 1+ e^{-r/\xi} r^{d-2}$.
%En 3D on peut citer \cite{Martin1987,Chen1986}

Figure~\ref{fig:acidification}c of the main text and Supplementary Figure~\ref{fig:nonmonotonic}c and d show the evolution of the fitting parameters in two different sets of boundary conditions: (i) complete adhesion of the gel on all walls, preventing syn\ae{}resis and swelling; (ii) complete adhesion except on the top wall, allowing syn\ae{}resis and swelling. The later situation is a proxy to the situation where we observe wrinkling, since it is necessary to keep the gel into the thin focal plane of the 63x objective lens of the confocal microscope. To do so, we coat only the microscope slide with acrylamide brush. The cover slip is cleaned but not coated, thus keeping strong adhesion on the side closer to the objective lens. Neither buckling nor wrinkling is detected in such configuration.

In situation (i), the gel keeps its topology throughout the gelation and over acidification. Left of Supplementary Figure~\ref{fig:nonmonotonic}a shows that the arms of the gel first shrink laterally while keeping their length, and then swell again. Syn\ae{}resis and swelling are thus local processes. By contrast in situation (ii), the gel as a whole compresses while forming. Since our focal plane is a fixed depth, we cannot follow the topology of the same subset of the network, however the pore size is clearly smaller than in situation (ii). The isoelectric gel, at minimum gel volume, is probably isotropic or only marginally anisotropic in $z$. This explains that subsequent swelling is not occurring only along $z$ but also in plane, causing buckling.

@@

\begin{figure*}%
	\includegraphics{permeability}%
	\caption{Permeability measurements. (a) Schematic representation of the experiment. (b-c) Evolution of the height of the interface in tube 1 relative to the final height in tube 2. The black line is the best exponential fit $Ae^{-t/\tau}$. (b) Gel is 4\%w casein, 4\%w GDL in water, $H=\SI{2.3}{\milli\metre}$ and $\tau=\SI{57}{\minute}$. (c) Same as (b) for a 50\%w glycerol--water mixture, $H=\SI{4}{\milli\metre}$ and $\tau=\SI{100}{\hour}$.}%
	\label{fig:permeability}%
\end{figure*}
\subsection*{Permeability measurements}

@@

By fitting $z(t)-z_\infty$ to an exponential, see Supplementary Figure~\ref{fig:permeability} b-c, we obtain $\tau$ and thus $\alpha$. Note that even with $H$ of the order of a few millimetres high glycerol contents yield $\tau\approx \SI{100}{\hour}$ due both to high viscosity and low permeability. Obtaining $z_\infty$ without tube 2 would be time prohibitive.

@@

From these data, we obtain for each stack the volume and thickness of the gel phase, area of the mid surface and peak-to-peak amplitude. Velocity is obtained by time differentiation. Once the instability has saturated on the top and bottom walls of the cell, peaks in the histograms of $z_\text{top}$ and $z_\text{bottom}$ yield the measure of the positions of both walls, thus of the cell thickness $e$ and retrospectively of $H_1$ and $H_2$.

\subsection*{Characteristic time measurement}
\begin{figure}
	\includegraphics{growth}
	\caption{Growth of the instability. (a) Evolution of the gel velocity $v$ averaged over the whole surface. Inset: Evolution of the peak-to-peak amplitude $A(t)$ in semi logarithmic scale. The black line is the best exponential fit with $\tau=\SI{90}{\second}$. (b) Evolution of the bottom profile of the gel from onset of instability (yellow) to contact to both walls (brown). Note that the longitudinal size of the blister evolves in this regime. The thick black curve is the profile at much longer times and is at lower altitude due to swelling. Note that the blister foot, highlighted by the orange vertical line, has not moved since contact.}
	\label{fig:growth}
\end{figure}

@@

We decomposed the deflection in sinusoidal modes $w(x,t) =  A \sin qx$ of wavevector $q$. Equation~(\ref{eq:mecheq}) then reads
\begin{equation}
\left(\frac{1}{12} B h^3 q^4 - \sigma_\parallel h q^2\right)w = \sigma_\perp,
\label{eq:mecheqop}
\end{equation}
where $B=E/(1-\nu^2)$ is the modulus of the plate.

@@

\begin{figure}
\includegraphics{notations}%
\caption{Schematic side view of the slit. The constant thickness gel film (orange) is surrounded by solvent. The region of interest of the mass conservation is highlighted in gray. Arrows are oriented as x or z to allow algebraic notations.}%
\label{fig:notations}%
\end{figure}

Let us now turn to the case of an elastic, impermeable film on a single viscous layer. As \cite{Huang2002} we neglect gravity, although it is difficult to do so in practice at a free interface. Assuming the layer thickness $H$ to be much smaller than $\lambda_d$, we can use the lubrication approximation. Further considering only small deflections, we may neglect the displacement of the plate along $x$. The flux along $x$ is thus of the Poiseuille form
\begin{equation}
Q = -\frac{H^3}{12\eta}\frac{\partial p}{\partial x}.
\label{eq:PoiseuilleFlux}
\end{equation}
where $p(x)$ is the pressure in the layer of viscosity $\eta$.

@@

Combining Equations~(\ref{eq:mecheqop}) and (\ref{eq:PoiseuilleLoad}), the compression load writes
\begin{equation}
\sigma_\parallel = \frac{1}{12}B h^2 q^2 + 12\frac{\eta}{\tau}\frac{1}{H^3 h q^4}
\label{eq:sigma0P}
\end{equation}
and is minimised at $q_P$ such that
\begin{equation}
q_P^6 = 2\times 12^2 \frac{1}{h^3H^3\Upsilon},
\label{eq:qP}
\end{equation}
with $\Upsilon = B\tau/\eta$ the viscoelastic factor. The dominant wavelength for the Poiseuille model is thus
\begin{equation}
\lambda_P = \pi\sqrt{hH}\left(\frac{2}{9}\Upsilon\right)^{1/6}
\end{equation}

We note that the mode minimising $\sigma_\parallel$ is also maximising the growth rate
\begin{equation}
\frac{1}{\tau} = \frac{H^3q^2}{12^2\eta}\left(12\sigma_\parallel - Bh^3q^4\right).
\end{equation}

When the elastic film is sandwiched between two viscous layers of respective thickness $H_1$ and $H_2$, the linearity of equation~(\ref{eq:mecheq}) allows us to use an effective substrate of thickness 
\begin{equation}
\frac{1}{H^3} = \frac{1}{H_1^3}+\frac{1}{H_2^3}.
\end{equation}
This corresponds to the experimental case where the gel film is far from both of the cell walls. Since we have an almost buoyant gel and no more free surface gravity can indeed be neglected.

@@

Combining equations~(\ref{eq:mecheqop}) and (\ref{eq:DarcyLoad}), the compression load writes
\begin{equation}
\sigma_\parallel = \frac{1}{12}B h^2 q^2 + \frac{\eta}{\tau}\frac{1}{\alpha q^2}
\label{eq:sigma0D}
\end{equation}
and is minimised at $q_D$ such that
\begin{equation}
q_D^4 = \frac{12}{h^2\alpha\Upsilon}.
\label{eq:qD}
\end{equation}

The dominant wavelength for Darcy model is thus
\begin{equation}
\lambda_D = 2\pi h^{1/2}\alpha^{1/4}\left(\frac{\Upsilon}{12}\right)^{1/4}.
\end{equation}
Here also the dominant mode maximises the growth rate.

@@

The mass conservation in the lower viscous layer is the same as equation~(\ref{eq:conservation}) with an added leak $v$ due to the porosity
\begin{equation}
\frac{\partial Q_1}{\partial x} + \frac{\partial w}{\partial t} + v = 0.
\label{eq:conservationDarcy}
\end{equation}

$v$ can be expressed by the Darcy law
\begin{equation}
v = \frac{\alpha}{\eta} \frac{p_1-p_2}{h}.
\label{eq:Darcy}
\end{equation}

Combining equations~(\ref{eq:pressures}), (\ref{eq:conservationDarcy}) and (\ref{eq:Darcy}) we obtain the transverse load
\begin{equation}
\sigma_\perp = p_1-p_2 = - \frac{\eta}{\tau} \frac{w}{\frac{H^3 q^2}{12} + \frac{\alpha}{h}},
\label{eq:MixtLoad}
\end{equation}

Note that we recover equation~(\ref{eq:PoiseuilleLoad}) when $\alpha \rightarrow 0$ and equation~(\ref{eq:DarcyLoad}) when $H \rightarrow 0$. We can recast this intuition in terms of the dimensionless number $H/H^*$ where $H^*$ is obtained by equating $q_P$ and $q_D$:
\begin{equation}
H^* = 2^{2/3} 3^{1/6} \alpha^{1/2} \Upsilon^{1/6}.
\end{equation}

@@

\subsection*{Samples characteristics}
To test the theoretical predictions we have fully characterised 9 samples, see Supplementary Table \ref{tab:data}.

\begin{table*}
\begin{tabular}{cSSSScSrrScSSSScSSSSSS}
%\toprule
&\multicolumn{4}{c}{Preparation} & &  \multicolumn{4}{c}{Properties} & & \multicolumn{4}{c}{Measurements} & & \multicolumn{6}{c}{Results} \\ 
\cline{2-5} \cline{7-10} \cline{12-15} \cline{17-22}\\[-2ex]
\# &{cas} & {GDL} & {gly} & {$e$} && {$\eta$} & \multicolumn{1}{c}{$G^\prime$} & \multicolumn{1}{c}{$\alpha$} & {$\xi$} && {$h$} & {$H_1$} & {$H_2$} & {$v_\text{max}$} && {$\Upsilon$} & {$H/H^*$} & {$\lambda_\text{exp}$} & {$\lambda_{D}$} & {$\lambda_{P}$} & {$\lambda_{D+P}$} \\ 
&{\%w} & {\%w} & {\%w} & \si{\micro\metre} && \si{\milli\pascal\second} & \multicolumn{1}{c}{\si{\pascal}} & \si{\square\nano\metre} & \si{\micro\metre} && \si{\micro\metre} & \si{\micro\metre} & \si{\micro\metre} & \si{\micro\metre/\second} && {$10^8$} &  & \si{\milli\metre} & \si{\milli\metre} & \si{\milli\metre} & \si{\milli\metre} \\ 
\cline{2-5} \cline{7-10} \cline{12-15} \cline{17-22}\\[-2ex]
%ech2
a & 4 & 4 & 0 & 74.5 && 1.0 & 487 & 67000 & 4.9 && 38.3 & 6.3 & 29.3 & 0.11 && 1.62 & 0.54 & 0.89 & 1.20 & 0.88 & 1.22 \\ 
& & & & && & & & && 40.2 & 1.7 & 32.9 & 0.09 && 0.70 & 0.09 & 0.45 & 1.00 & 0.39 & 1.00\\
b & 4 & 4 & 0 & 92.4 && 1.0 & 487 & 67000 & 4.9 && 27.8 & 13.5 & 50.8 & 0.20 && 1.57 & 1.17 & 0.85 & 1.01 & 1.10 & 1.20 \\ 
c & 4 & 4 & 0 & 106.1 && 1.0 & 487 & 67000 & 4.9 && 32.7 & 30.7 & 42.9 & 0.14 && 2.54 & 2.23 & 1.52 & 1.24 & 1.85 & 1.90 \\ 
d & 4 & 4 & 0 & 138.0 && 1.0 & 487 & 67000 & 4.9 && 48.2 & 32.0 & 63.9 & 0.11 && 4.05 & 2.28 & 1.62 & 1.69 & 2.56 & 2.62 \\
& & & & && & & & && 57.4 & 2.1 & 80.9 & 0.05 && 1.25 & 0.09 & 0.81 & 1.37 & 0.6 & 1.37 \\
\cline{2-5} \cline{7-10} \cline{12-15} \cline{17-22}\\[-2ex]
e & 8 & 8 & 0 & 85.2 && 1.0 & 1817 & 2150 & 3.1 && 49.8 & 10.0 & 26.8 & 0.10 && 6.50 & 3.79 & 1.00 & 0.82 & 1.60 & 1.61 \\ 
& & & & && & & & && 60.5 & 1.1 & 24.0 & 0.11 && 1.94 & 0.42 & 0.50 & 0.77 & 0.55 & 0.82\\
f & 8 & 8 & 0 & 91.0 && 1.0 & 1817 & 2150 & 3.1 && 23.4 & 30.2 & 47.7 & 0.10 && 12.05 & 9.70 & 1.50 & 0.66 & 2.04 & 2.04 \\ 
& & & & && & & & && 30.7 & 1.6 & 55.0 & 0.02 && 15.60 & 0.32 & 0.75 & 0.82 & 0.59 & 0.83 \\
\cline{2-5} \cline{7-10} \cline{12-15} \cline{17-22}\\[-2ex]
g & 4 & 4 & 20 & 63.7 && 1.6 & 1142 & 24600 & 2.7 && 39.1 & 1.2 & 10.9 & 0.21 && 0.86 & 0.20 & 0.49 & 0.80 & 0.36 & 0.80 \\ 
h & 4 & 4 & 35 & 82.3 && 2.7 & 1858 & 25300 & 1.8 && 58.3 & 0.9 & 12.6 & 0.02 && 10.98 & 0.13 & 0.61 & 1.06 & 0.56 & 1.06 \\ 
i & 4 & 4 & 50 & 95.3 && 5.4 & 4660 & 5700 & 1.5 && 65.1 & 7.9 & 14.4 & 0.02 && 16.30 & 1.52 & 1.23 & 1.50 & 1.85 & 1.96 \\ 
\end{tabular}
\caption{Fully characterised samples at the basis of Figure~\ref{fig:DarcyPoiseuille}. Lines where preparation and properties are left blank correspond to the average of the secondary blisters of the previous line. `cas', `GDL' and `gly' indicate the weight fraction of sodium caseinate, GDL and glycerol respectively. $\xi$ is the pore size at the end of each experiment. $h$ is the thickness of the gel film just before buckling.}
\label{tab:data}
\end{table*}

>>>>

\paragraph*{Acknowledgements}
Confocal experiments were conducted at SFR BioSciences Gerland - Lyon Sud (US8 / UMS3444).
The authors thank Jos\'e Bico, Arezeki Boudaoud, Cyprien Gay, L. Mahadevan and Olivier Pierre-Louis for theoretical insights at various stages of the research. 
Precious experimental ideas came from Thibaut Divoux, Catherine Barentin, Denis Bartolo, Elisabeth Bouchaud and Maxime Lefranc.
Alan Parker at Firmenich provided the casein and GDL; 
Special thanks to Madame Biot who made M.A. Biot work accessible through the \href{http://www.olemiss.edu/sciencenet/poronet/}{poronet} website. 
ML thanks the Region Rhône Alpes and the Programme d'Avenir Lyon - Saint Etienne (PALSE NoGELPo) for postoctoral grant. 
TG and MN acknowledge funding from the Agence Nationale de la Recherche (ANR-11-PDOC-027). 
SM and ML acknowledge funding from the European Research Council under the European Union's Seventh Framework Program (FP7/2007-2013) / ERC grant agreement No. 258803 and from Institut Universitaire de France.

@@

\subsection*{Fraction of free casein as a function of pH}

The fraction of free casein proteins of a \SI{100}{\milli\litre}, 4\%w sodium caseinate dispersion is obtained either by titration by 1 molar HCl, see Supplementary Video~\ref{vid:titration}, or during the action of 4\%w GDL. In both cases, \SI{0.4}{\milli\litre} of the \SI{100}{\milli\litre} dispersion is sampled regularly as a function of pH, and centrifuged (centrifuge 5424, Eppendorf) in Eppendorf tubes at 14500 rpm for \SI{3}{\minute}. The supernatant is diluted and analysed with a UV absorption spectrometer (USB4000, Ocean Optics). According to~\cite{Roefs1986}, we deduce the fraction of free casein proteins from the adsorption spectrum as:
%
\begin{align}
x_\text{free} &= \frac{A(\text{pH})}{A(\text{pH}=7)},\\
\text{with }A(\text{pH}) &= A_{280}-1.706 A_{320},
\end{align}
%
where $A_{280}$ is the absorption peak at \SI{280}{\nano\metre} due to aromatic amino acids in the casein molecules plus a turbidity background whereas the absorption peak at \SI{320}{\nano\metre} is only due to turbidity.

Above the isoelectric point, the homogeneous acidification by GDL continuously creates very small protein aggregates that are difficult to centrifuge. Such suspensions are too turbid and too rapidly evolving to reliably measure $x_\text{free}$.

@@

\begin{figure}
	\includegraphics{nonmonotonic}%
	\caption{Microstructure evolution of a 4\%w sodium caseinate solution acidified with 4\%w GDL in water. (a) Details of confocal images in fully adhesive conditions (top row) and with no adhesion on top (bottom row). The scale bar is \SI{10}{\micro\metre}. Arrows indicates the pore size. (b) Corresponding Fourier spectra when no adhesion on top, same times as (a). Lines are fits to equation~(\ref{eq:fractalS}). The inset shows a schematic association between real space structures and wave numbers. (c-e) Evolution of cut-off length $\xi$, susceptibility $\chi$ and fractal dimension $d$ in full adhesive (black line) and no adhesion on top (orange dashed line) situations. Symbols indicate the times of the images in (a).}%
	\label{fig:nonmonotonic}%
\end{figure}

To follow the evolution of the microstructure of the gel, we acquire slices of $(\SI{2048}{px})^2$ with pixels of \SI{98}{\nano\metre}, Supplementary Figure~\ref{fig:nonmonotonic}a and Supplementary Videos~\ref{vid:stick63} and \ref{vid:nostick63}. After Hann windowing we perform Fourier transform and radially average the resulting spectrum to obtain $I(q)$, see Supplementary Figure~\ref{fig:nonmonotonic}b that is fitted by the following fractal form~\cite{VanDijk1986}:
\begin{equation}
I(q) = \frac{2\chi\Gamma(d)}{\left(1+\left((d+1)\xi q\right)^2\right)^{d/2}}
\label{eq:fractalS}
\end{equation}
where the susceptibility $\chi$ is related to the fluorescence contrast between the network and the solvent, the cut-off length $\xi$ corresponds to the largest pore size and $d$ is the fractal dimension. $\Gamma$ is the Gamma function. Equation~(\ref{eq:fractalS}) is a good approximation to the 2D Fourier transform of a fractal pair correlation function $g(r) \sim 1+ e^{-r/\xi} r^{d-2}$.
%En 3D on peut citer \cite{Martin1987,Chen1986}

Figure~\ref{fig:acidification}d of the main text and Supplementary Figure~\ref{fig:nonmonotonic}c-e show the evolution of the fitting parameters in two different sets of boundary conditions: (i) complete adhesion of the gel on all walls, preventing syn\ae{}resis and swelling; (ii) complete adhesion except on the top wall, allowing syn\ae{}resis and swelling. The latter situation is a proxy to the situation where we observe wrinkling, since it is necessary to keep the gel into the thin focal plane of the 63x objective lens of the confocal microscope. To do so, we coat only the microscope slide with acrylamide brush. The cover slip is cleaned but not coated, thus keeping strong adhesion on the side closer to the objective lens. Neither buckling nor wrinkling is detected in such a configuration.

In situation (i), the gel keeps its topology throughout the gelation and over-acidification. Left of Supplementary Figure~\ref{fig:nonmonotonic}a shows that the arms of the gel first shrink laterally while keeping their length, and then swell again. Syn\ae{}resis and swelling are thus local processes. By contrast in situation (ii), the gel as a whole compresses while forming. Since our focal plane is at a fixed depth, we cannot follow the topology of the same subset of the network, however the pore size is clearly smaller than in situation (ii). The isoelectric gel, at minimum gel volume, is probably isotropic or only marginally anisotropic in $z$. This explains that subsequent swelling is not occurring only along $z$ but also in plane, causing buckling.

@@

\begin{figure*}%
	\includegraphics{permeability}%
	\caption{Permeability measurements. (a) Schematic representation of the experimental set up. (b-c) Evolution of the height of the interface in tube 1 relative to the final height in tube 2. The black line is the best exponential fit $Ae^{-t/\tau}$. (b) Gel is 4\%w casein, 4\%w GDL in water, $H=\SI{2.3}{\milli\metre}$ and $\tau=\SI{57}{\minute}$. (c) Same as (b) for a 50\%w glycerol--water mixture, $H=\SI{4}{\milli\metre}$ and $\tau=\SI{100}{\hour}$.}%
	\label{fig:permeability}%
\end{figure*}
\subsection*{Permeability measurements}

@@

By fitting $z(t)-z_\infty$ to an exponential, see Supplementary Figure~\ref{fig:permeability} b-c, we obtain $\tau$ and thus $\alpha$. Note that even with $H$ of the order of a few millimetres high glycerol contents yield $\tau\approx \SI{100}{\hour}$ due to both high viscosity and low permeability. Obtaining $z_\infty$ without tube 2 would be time prohibitive.

@@

From these data, we obtain for each stack the volume and thickness of the gel phase, area and peak-to-peak amplitude of the mid surface. Velocity is obtained by time differentiation. Once the instability has saturated on the top and bottom walls of the cell, peaks in the histograms of $z_\text{top}$ and $z_\text{bottom}$ yield the measure of the positions of both walls, thus of the cell thickness $e$ and retrospectively of $H_1$ and $H_2$.

\subsection*{Characteristic time measurement}
\begin{figure}
	\includegraphics{growth}
	\caption{Growth of the instability. (a) Evolution of the gel velocity $v$ averaged over the whole surface. Inset: Evolution of the peak-to-peak amplitude $A(t)$ in semi logarithmic scales. The black line is the best exponential fit with $\tau=\SI{90}{\second}$. (b) Evolution of the bottom profile of the gel from onset of instability (yellow) to contact to both walls (brown). Note that the longitudinal size of the blister evolves in this regime. The thick black curve is the profile at much longer times and is at lower altitude due to swelling. Note that the blister foot, highlighted by the orange vertical line, has not moved since contact.}
	\label{fig:growth}
\end{figure}

@@

We decompose the deflection in sinusoidal modes $w(x,t) =  A \sin (qx)$ of wavevector $q$. Equation~(\ref{eq:mecheq}) then reads
\begin{equation}
\left(\frac{1}{12} B h^3 q^4 - \sigma_\parallel h q^2\right)w = \sigma_\perp,
\label{eq:mecheqop}
\end{equation}
where $B=E/(1-\nu^2)$ is the modulus of the plate.

@@

\begin{figure}
\includegraphics{notations}%
\caption{Schematic side view of the slit. The constant thickness gel film (orange) is surrounded by solvent. The region of interest for mass conservation is highlighted in gray. Arrows are oriented along $x$ or $z$, defining positive sign.}%
\label{fig:notations}%
\end{figure}

Let us now turn to the case of an elastic, impermeable film on a single viscous layer. As in \cite{Huang2002} we neglect gravity, although it would be difficult to do so in practice at a free interface. Assuming the layer thickness $H$ to be much smaller than $\lambda_d$, we can use the lubrication approximation. Further considering only small deflections, we may neglect the displacement of the plate along $x$. The flux along $x$ is thus of the Poiseuille form
\begin{equation}
Q = -\frac{H^3}{12\eta}\frac{\partial p}{\partial x}.
\label{eq:PoiseuilleFlux}
\end{equation}
where $p(x)$ is the pressure in the layer of viscosity $\eta$.

@@

Combining Equations~(\ref{eq:mecheqop}) and (\ref{eq:PoiseuilleLoad}), the compression load writes
\begin{equation}
\sigma_\parallel = \frac{1}{12}B h^2 q^2 + 12\frac{\eta}{\tau}\frac{1}{H^3 h q^4}
\label{eq:sigma0P}
\end{equation}
and is minimised at $q_P$ such that
\begin{equation}
q_P^6 = 2\times 12^2 \frac{1}{h^3H^3\Upsilon},
\label{eq:qP}
\end{equation}
with $\Upsilon = B\tau/\eta$ the viscoelastic factor. The dominant wavelength for the Poiseuille model is thus
\begin{equation}
\lambda_P = \pi\sqrt{hH}\left(\frac{2}{9}\Upsilon\right)^{1/6}.
\end{equation}
%
We note that the mode minimising $\sigma_\parallel$ is also maximising the growth rate
\begin{equation}
\frac{1}{\tau} = \frac{H^3q^2}{12^2\eta}\left(12\sigma_\parallel - Bh^3q^4\right).
\end{equation}

When the elastic film is sandwiched between two viscous layers of respective thickness $H_1$ and $H_2$, the linearity of equation~(\ref{eq:mecheq}) allows us to use an effective substrate of thickness 
\begin{equation}
\frac{1}{H^3} = \frac{1}{H_1^3}+\frac{1}{H_2^3}.
\end{equation}
This corresponds to the experimental case where the gel film is far from both of the cell walls. Since we have an almost buoyant gel and no more free surface, gravity can indeed be neglected.

@@

Combining equations~(\ref{eq:mecheqop}) and (\ref{eq:DarcyLoad}), the compression load writes
\begin{equation}
\sigma_\parallel = \frac{1}{12}B h^2 q^2 + \frac{\eta}{\tau}\frac{1}{\alpha q^2}
\label{eq:sigma0D}
\end{equation}
and is minimised at $q_D$ such that
\begin{equation}
q_D^4 = \frac{12}{h^2\alpha\Upsilon}.
\label{eq:qD}
\end{equation}
%
The dominant wavelength for Darcy model is thus
\begin{equation}
\lambda_D = 2\pi h^{1/2}\alpha^{1/4}\left(\frac{\Upsilon}{12}\right)^{1/4}.
\end{equation}
Here also the dominant mode maximises the growth rate.

@@

The mass conservation in the lower viscous layer is the same as equation~(\ref{eq:conservation}) with an added leak $v$ due to the porosity
\begin{equation}
\frac{\partial Q_1}{\partial x} + \frac{\partial w}{\partial t} + v = 0.
\label{eq:conservationDarcy}
\end{equation}
%
$v$ can be expressed by the Darcy law
\begin{equation}
v = \frac{\alpha}{\eta} \frac{p_1-p_2}{h}.
\label{eq:Darcy}
\end{equation}

Combining equations~(\ref{eq:pressures}), (\ref{eq:conservationDarcy}) and (\ref{eq:Darcy}) we obtain the transverse load
\begin{equation}
\sigma_\perp = p_1-p_2 = - \frac{\eta}{\tau} \frac{w}{\frac{H^3 q^2}{12} + \frac{\alpha}{h}},
\label{eq:MixtLoad}
\end{equation}
%
Note that we recover equation~(\ref{eq:PoiseuilleLoad}) when $\alpha \rightarrow 0$ and equation~(\ref{eq:DarcyLoad}) when $H \rightarrow 0$. We can recast this intuition in terms of the dimensionless number $H/H^*$ where $H^*$ is obtained by equating $q_P$ and $q_D$:
\begin{equation}
H^* = 2^{2/3} 3^{1/6} \alpha^{1/2} \Upsilon^{1/6}.
\end{equation}

@@

\subsection*{Samples characteristics}
To test the theoretical predictions we have fully characterised 9 samples. Their characteristics are gathered in Supplementary Table \ref{tab:data}.

\begin{table*}
\begin{tabular}{cSSSScSrrScSSSScSSSSSS}
%\toprule
&\multicolumn{4}{c}{Preparation} & &  \multicolumn{4}{c}{Properties} & & \multicolumn{4}{c}{Measurements} & & \multicolumn{6}{c}{Results} \\ 
\cline{2-5} \cline{7-10} \cline{12-15} \cline{17-22}\\[-2ex]
\# &{cas} & {GDL} & {gly} & {$e$} && {$\eta$} & \multicolumn{1}{c}{$G^\prime$} & \multicolumn{1}{c}{$\alpha$} & {$\xi$} && {$h$} & {$H_1$} & {$H_2$} & {$v_\text{max}$} && {$\Upsilon$} & {$H/H^*$} & {$\lambda_\text{exp}$} & {$\lambda_{D}$} & {$\lambda_{P}$} & {$\lambda_{D+P}$} \\ 
&{\%w} & {\%w} & {\%w} & \si{\micro\metre} && \si{\milli\pascal\second} & \multicolumn{1}{c}{\si{\pascal}} & \si{\square\nano\metre} & \si{\micro\metre} && \si{\micro\metre} & \si{\micro\metre} & \si{\micro\metre} & \si{\micro\metre/\second} && {$10^8$} &  & \si{\milli\metre} & \si{\milli\metre} & \si{\milli\metre} & \si{\milli\metre} \\ 
\cline{2-5} \cline{7-10} \cline{12-15} \cline{17-22}\\[-2ex]
%ech2
a & 4 & 4 & 0 & 74.5 && 1.0 & 487 & 67000 & 4.9 && 38.3 & 6.3 & 29.3 & 0.11 && 1.62 & 0.54 & 0.89 & 1.20 & 0.88 & 1.22 \\ 
& & & & && & & & && 40.2 & 1.7 & 32.9 & 0.09 && 0.70 & 0.09 & 0.45 & 1.00 & 0.39 & 1.00\\
b & 4 & 4 & 0 & 92.4 && 1.0 & 487 & 67000 & 4.9 && 27.8 & 13.5 & 50.8 & 0.20 && 1.57 & 1.17 & 0.85 & 1.01 & 1.10 & 1.20 \\ 
c & 4 & 4 & 0 & 106.1 && 1.0 & 487 & 67000 & 4.9 && 32.7 & 30.7 & 42.9 & 0.14 && 2.54 & 2.23 & 1.52 & 1.24 & 1.85 & 1.90 \\ 
d & 4 & 4 & 0 & 138.0 && 1.0 & 487 & 67000 & 4.9 && 48.2 & 32.0 & 63.9 & 0.11 && 4.05 & 2.28 & 1.62 & 1.69 & 2.56 & 2.62 \\
& & & & && & & & && 57.4 & 2.1 & 80.9 & 0.05 && 1.25 & 0.09 & 0.81 & 1.37 & 0.6 & 1.37 \\
\cline{2-5} \cline{7-10} \cline{12-15} \cline{17-22}\\[-2ex]
e & 8 & 8 & 0 & 85.2 && 1.0 & 1817 & 2150 & 3.1 && 49.8 & 10.0 & 26.8 & 0.10 && 6.50 & 3.79 & 1.00 & 0.82 & 1.60 & 1.61 \\ 
& & & & && & & & && 60.5 & 1.1 & 24.0 & 0.11 && 1.94 & 0.42 & 0.50 & 0.77 & 0.55 & 0.82\\
f & 8 & 8 & 0 & 91.0 && 1.0 & 1817 & 2150 & 3.1 && 23.4 & 30.2 & 47.7 & 0.10 && 12.05 & 9.70 & 1.50 & 0.66 & 2.04 & 2.04 \\ 
& & & & && & & & && 30.7 & 1.6 & 55.0 & 0.02 && 15.60 & 0.32 & 0.75 & 0.82 & 0.59 & 0.83 \\
\cline{2-5} \cline{7-10} \cline{12-15} \cline{17-22}\\[-2ex]
g & 4 & 4 & 20 & 63.7 && 1.6 & 1142 & 24600 & 2.7 && 39.1 & 1.2 & 10.9 & 0.21 && 0.86 & 0.20 & 0.49 & 0.80 & 0.36 & 0.80 \\ 
h & 4 & 4 & 35 & 82.3 && 2.7 & 1858 & 25300 & 1.8 && 58.3 & 0.9 & 12.6 & 0.02 && 10.98 & 0.13 & 0.61 & 1.06 & 0.56 & 1.06 \\ 
i & 4 & 4 & 50 & 95.3 && 5.4 & 4660 & 5700 & 1.5 && 65.1 & 7.9 & 14.4 & 0.02 && 16.30 & 1.52 & 1.23 & 1.50 & 1.85 & 1.96 \\ 
\end{tabular}
\caption{Characteristics of the samples used for Figure~\ref{fig:DarcyPoiseuille}. Lines where preparation and properties are left blank correspond to the average of the secondary blisters of the previous line. `cas', `GDL' and `gly' indicate the weight fraction of sodium caseinate, GDL and glycerol respectively. $\xi$ is the pore size at the end of each experiment. $h$ is the thickness of the gel film just before buckling.}
\label{tab:data}
\end{table*}
